---
title: "R Code Section 7 - Discrete-time, discrete-state multistate Markov models from the perspective of algebraic statistics"
author: "Etienne Sebag"
date: "2026-01-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## AI Disclosure: 

Generative AI tools were used in generating smaller sections of the code that follows, mostly for the purposes of quick and efficient data transformation and to assist in smaller coding tasks such as de-bugging and safety checks. 


## Preliminaries. 

Let us recall the essentials: 

A) We have 27 states (26 states (which are the letters)  plus one absorbing state (the space). 

B) Each word corresponds to a specific trajectory, or path. 

C) **We want each trajectory to have the same length.** Hence, we are padding the trajectories with a "space", and considering that the maximum length L is the longest word in the state space. So the time horizon of the process goes from $l=1,2...,L, L+1.$

D) Each location of a letter is a time instance/time point. According to the notations in the overleaf,the time index starts from 1 so we have $X_1, X_2, ..., X_L,X_{L+1}$. (in our case).

In the example of "house" word,
{X1 = h, X2 = o, X3 = u, X4 = s, X5 = e}. If L = 8 then we will add X6
= X7 = X8 = 27 (absorbing state, space.) [other letters should be numbered forthe sake of consistency]. Each of such sequence contributes to the
likelihood.


## Importing the Data

```{r}
# import
data = "shakespeare.txt"

# making the file be essentially one big string
txt = paste(readLines(data, warn = FALSE, encoding = "UTF-8"), collapse = " ")
```

## Initial Cleaning

```{r}
# making everything lower case just for convenience sake and so we only deal with 26 states (a-z) + the 1 absorbing state (space)

txt = tolower(txt)

# keeping only letters and spaces so that, basically, anything not a-z becomes a space
txt = gsub("[^a-z ]", " ", txt)

# collapsing multiple spaces into just a single space
txt = gsub(" +", " ", txt)

# trim leading/trailing spaces
txt = trimws(txt)
```

## Splitting Into Words

```{r}
words = strsplit(txt, " ", fixed = TRUE)[[1]]

# safecheck: drop any empty strings
words = words[words != ""]

#sanity check to see what is happening at this stage:

head(words,20)
```
##  Removing one-letter words

```{r}
words = words[nchar(words) > 1]

```



#  Fixing and working with the length of a word/trajectory:


```{r}

# If L is the maximal length of a word, we should have 1,2, ..., L, L+1 times just to have space for a space " " in position (L+1). 


lens = nchar(words)
u_lens = sort(unique(lens), decreasing = TRUE)

if (length(u_lens) < 2) stop("Not enough distinct word lengths to define a second-longest length.")

L = u_lens[2]        # extracting the second-longest word length just for visualization purposes
T_horizon = L + 1    # according to our notation and problem set-up in the manuscript

# WE Drop words longer than L (this removes the single longest and any other outliers)
words <- words[nchar(words) <= L]

# checks:

cat("Second-longest word length L =", L, "\n")
cat("Trajectory length (columns) L+1 =", T_horizon, "\n")
cat("Number of remaining word tokens =", length(words), "\n\n")
```

#  Building the array

GOAL:  Forming data that corresponds on an array of letters of equal length where after the 1st empty space we complete the array by adding (padding) mode empty spaces.

```{r}
# We recall: L is the maximum retained word length
# T_horizon is the number of time points, including the final space
T_horizon <- L + 1

make_traj <- function(w, T_horizon) {
  k <- nchar(w)
  paste0(w, strrep(" ", T_horizon - k))
}

traj <- vapply(words, make_traj, FUN.VALUE = character(1), T_horizon = T_horizon)
```

# Counting the trajectories

```{r}

tab = table(traj)                 # named integer vector
traj_unique = names(tab)
counts = as.integer(tab)

cat("Number of distinct trajectories =", length(traj_unique), "\n\n")
```

# Combining everything

```{r}

# turning distinct trajectory strings into a matrix (L+1 columns)
    # and adding/appending the count column at the end.

# and 
# Splitting each trajectory string into single characters
traj_list = strsplit(traj_unique, "", fixed = TRUE)
W_unique = do.call(rbind, traj_list)

# Safety check: should be exactly T_horizon columns
stopifnot(ncol(W_unique) == T_horizon)

# This (df) is the Final data frame we create after applying many cleaning steps and pre-processing steps:L letter columns + count
df <- as.data.frame(W_unique, stringsAsFactors = FALSE)
names(df) <- paste0("t", 1:L)   # FIX: only 1:L
df$count <- counts

```


```{r}
#Cleaning up "df": (column names), simply because before the "t_18" column did not have a name. 


names(df)[1:T_horizon] <- paste0("t", 1:T_horizon)

# keeping the count column name
names(df)[T_horizon + 1] <- "count"

# Check
names(df)


```


## Sanity and Verification checks: 

```{r}
words <- words[nchar(words) <= L]

# length of each retained word
lens <- nchar(words)

# and the longest retained length
max_len <- max(lens)

# Extracting words with that length
longest_words <- unique(words[lens == max_len])

max_len
longest_words
```
## Saving the file

```{r}
write.csv(df, "data_Shakespeare_final_github.csv", row.names = FALSE)
```


# Extra/Additional work for revisions of manuscript based on referee feedback: 

Computing the empirical frequency of "the" after applying the same cleaning and filtering steps that produce the above final dataset "df". 

```{r}
# Empirical frequency of "the" in the cleaned corpus represented by df

n_the <- sum(words == "the")
n_words <- length(words)

freq_the <- n_the / n_words

cat("Number of occurrences of 'the':", n_the, "\n")
cat("Total number of word tokens:", n_words, "\n")
cat("Empirical frequency of 'the':", freq_the, "\n")
```



###############################################################

## MLE computation for the Homogeneous case

```{r}
# setting-up

states = c(letters, " ") #this is basically the state space

S = length(states) # returns 27

print(S) #verifying
```


```{r}
t_cols <- df[, paste0("t", 1:T_horizon), drop = FALSE]
L_time <- ncol(t_cols)

#just to be safe:

if (L<2) stop("need to select and extract at least two time columns in order to make sense of a tpm")
```



```{r}
N_total  = sum(df$count) #total number of words INCLUDING their multiplicity

print(N_total)
```

## Working with the Initial Distribution

```{r}
#MLE FOR THE INITIAL DISTRIBUTION ROW VECTOR PI

start_counts = tapply(df$count, df$t1, sum)

# this counts how often (proportion) the first symbol (so at time t1) is a specific character (and weighted by count)

#but we need to normalize this to convert to probabilities

pi_hat = start_counts/N_total

print(pi_hat)
```



```{r}
#Based on Equation (6.6) in the manuscript. 

#recall: this quantity (the numerator) answers the question "how often did the process see/experience i-->j transition" in the first order (k=1) case:

#we first initialize a matrix of 27 by 27 and set all entries to 0. 

N_ij = matrix(0, nrow=S, ncol=S, dimnames = list(from=states, to =states))

#then we loop over the time steps and add these weighted transition counts

for (t in 1:(L_time-1)){
  
  from = t_cols[,t] #basically extracting the "i"'s
  to = t_cols[,t+1] #next time step, so extracting the "j"'s. 
  
 from_rep = rep(from, times = df$count) #counting
 to_rep = rep(to, times = df$count) #counting
 
 step_counts = table(factor(from_rep, levels=states),
 factor(to_rep, levels=states))
 
 N_ij = N_ij+step_counts
}
```

```{r}
#now dealing with N_i:

# N_i counts how many times state i appears as the current state
# before a transition to some next state j, including j = i.

N_i = rowSums(N_ij) #denominator for each row: total transitions starting from state i


```

```{r}
#Final computation:

# Normalizing each row to turn counts into probabilities
A_hat = N_ij
for (i in 1:S) {
  if (N_i[i] > 0) {
    A_hat[i, ] = N_ij[i, ] / N_i[i]
  } else {
    A_hat[i, ] = 0
  }
}
```


```{r}
print(pi_hat)
```
Preliminary conclusions: 

15% of words start with t. 
8% of words start with a. 
8% of words start with s. 

These observations can be added to the paper.  

Example of building a sub-matrix (based on common bigrams from the following website:) 

https://mathcenter.oxford.emory.edu/site/math125/englishLetterFreqs/


```{r}
#choosing our sub-matrix thoughtfully:
from_sel <- c("t", "h", "i", "e", "n") #from most common bigrams
to_sel   <- c("h", "e", "n", "s", "t")

A_5x5 <- A_hat[from_sel, to_sel] #sub-matrix to be displayed in the paper
round(A_5x5, 3)


```

##########################################################################

## MLE computation for the Non-Homogeneous Case:


```{r}
# initializing and creating empty containers
N_list = vector("list", L - 1)   # the counts per time
A_list = vector("list", L - 1)   #the  probs per time

# Looping over time steps
for (t in 1:(L_time - 1)) {

  # again, obtaining the from/to symbols at this time step
  from = df[[paste0("t", t)]]
  to   = df[[paste0("t", t + 1)]]

  # replicating each row count times 
  from_rep = rep(from, times = df$count)
  to_rep   = rep(to,   times = df$count)

  # counting transitions i->j at this time step using table
  N_t = table(
    factor(from_rep, levels = states),
    factor(to_rep,   levels = states)
  )

  # converting to numeric matrix
  N_t = as.matrix(N_t)

  # normalizing the rows to get probabilities
  row_sums = rowSums(N_t)
  A_t = N_t
  for (i in 1:S) {
    if (row_sums[i] > 0) A_t[i, ] = N_t[i, ] / row_sums[i]
    else A_t[i, ] = 0
  }

  # storing at each iteration
  N_list[[t]] = N_t
  A_list[[t]] = A_t

  cat("Done t =", t, "->", t + 1, "\n")
}

```
# Displaying the results in a useful way to include in the paper

```{r}
# choosing useful transitions to track (Table 4 in the paper)
pairs = list(
  c("t","h"),
  c("h","e"),
  c("i","n"),
  c("e","r"),
  c("a","l")
  
)

# Building a data frame
res = data.frame(
  t = 1:(L_time-1)
)

for (p in pairs) {
  lab = paste0(p[1], "->", ifelse(p[2]==" ","space",p[2]))
  res[[lab]] = sapply(A_list, function(A) A[p[1], p[2]])
}

round(res, 4)

```
 In this above table, the R output ``t=1`` is looking at transition pair/selected letter pair ($X_1 \rightarrow X_2$) (this corresponds to $l=2$ in the Table 4 of the manuscript). Similarly, position $13\rightarrow 14$ is designated by $l=14$ in Table 4, but by the row $t=13$ in this R code.  

####################################################################################################################################################

## Higher Order and Reversible-Illness Death Model


```{r}
#Mapping letters now to the reduced 3-state state space: (V,C,space)

vowels = c("a","e","i","o","u") #selecting the vowels

#creating a function

# x is a vector of characters we input to the function

# We will use S to designate a space in the new dataframe
to_VCspace = function(x) {
  
  out = ifelse(x == " ", "S",            # S for space (absorbing)
                ifelse(x %in% vowels, "V", "C"))
  out
}

states3 = c("V","C","S")

```

```{r}
# Extracting time columns
t_cols <- df[, paste0("t", 1:T_horizon), drop = FALSE]

# check: this should be a data frame
str(t_cols)

# Safety check: all selected columns should be character columns
sapply(t_cols, is.character)

# apply the mapping we made above to each time column so we are re-making our df

X <- as.data.frame(
  lapply(t_cols, to_VCspace), # each column gets passed to the function we made above
  stringsAsFactors = FALSE
)

# Keeping the same column names
names(X) <- names(t_cols)

# Safety checks
stopifnot(ncol(X) == T_horizon)
stopifnot(identical(names(X), names(t_cols)))

head(X)

```

## Saving the File: 

```{r}
write.csv(X, "consonant-vowels_final_github.csv", row.names=FALSE)
```

## Working with the Inference and Estimation 

```{r}
## 

contexts = c("VV","VC","CV","CC") #establishing the available pairs of states for the two previous time steps (ie: the previous two symbols). 

next_states = c("V","C","S") #what the next available state can be (it can be the whole reduced three-state state space $S^{*}$ according to Overleaf notation)

# count table: rows = pairs of the last two states (V,C), cols = next state
C_ctx <- matrix(
  0,
  nrow = length(contexts), #length of rows,
  ncol = length(next_states),#length of columns,
  dimnames = list(
    context = contexts,
    next_state = next_states
  )
)

#iterating:

for (t in 2:(L_time - 1)) {
  prev = X[[paste0("t", t - 1)]] #the letter at position l-2
  curr = X[[paste0("t", t)]] #the letter at position l-1
  nxt  = X[[paste0("t", t + 1)]] #the letter at position l

  # only contexts where prev and curr are in {V,C}
  
  
   #only interested in previous ones that are V or C
  
  keep = (prev != "S")  & (curr != "S")   # after S, we don't care (absorbing padding)

  if (!any(keep)) next
  
  #here we deal with the multiplicities:

  prev_rep = rep(prev[keep], times = df$count[keep])
  curr_rep = rep(curr[keep], times = df$count[keep])
  nxt_rep  = rep(nxt[keep],  times = df$count[keep])

  ctx_rep = paste0(prev_rep, curr_rep)   # "VV","VC","CV","CC"

  step_counts = table(factor(ctx_rep, levels = contexts),
                       factor(nxt_rep, levels = next_states))

  C_ctx = C_ctx + as.matrix(step_counts)
}

# row-normalize to get probabilities
rs = rowSums(C_ctx)
A2_ctx = C_ctx
for (r in 1:nrow(C_ctx)) {
  if (rs[r] > 0) A2_ctx[r,] = C_ctx[r,] / rs[r]
  else A2_ctx[r,] = 0
}

A2_ctx

```

## MLE of the Initial Distribution

```{r}

contexts = c("VV","VC","CV","CC")

x1 = X$t1
x2 = X$t2

# keep only starts where both are V or C (exclude any early S just in case)
keep12 = (x1 != "S") & (x2 != "S")

# expand by multiplicities 
x1_rep = rep(x1[keep12], times = df$count[keep12])
x2_rep = rep(x2[keep12], times = df$count[keep12])

ctx_rep = paste0(x1_rep, x2_rep)

# counts of starting contexts
M_ctx = table(factor(ctx_rep, levels = contexts))
M_ctx

# normalizing so that they are probabilities indeed
pi2_hat = as.numeric(M_ctx) / sum(M_ctx)
names(pi2_hat) = contexts
pi2_hat

```

## Addressing additional reviewer comments: 

Here, we want to take care of the following comment: 

*In light of the observation from Chapter 6, Example 4, the ordering of paths or words in the MLE may not match the ordering of paths or words in the observations. It would be interesting to identify a few outlier words in either direction in the dataset.*

```{r}
## Code chunk to observe and work with the words whose observed and MLE rankings may differ ## . 


# Recall: This diagnostic is based on our R object "df", which is the cleaned Shakespeare corpus.

# also: each row of df is one distinct Markov trajectory/word,

# and df$count gives the number of times that word occurs in the corpus.

# Making sure pi_hat has all 27 states, including zero entries if needed
pi_full <- setNames(rep(0, length(states)), states)
pi_full[names(pi_hat)] <- pi_hat

# a function to compute fitted path probability under the homogeneous MLE case
path_prob_homogeneous <- function(path, pi, A) {
  # path is a character vector, e.g. c("t","h","e"," "," ",...)
  
  p <- pi[path[1]]
  
  for (tt in 1:(length(path) - 1)) {
    p <- p * A[path[tt], path[tt + 1]]
  }
  
  return(as.numeric(p))
}

# Extracting the trajectory columns t1, ..., t_{L+1}
t_cols <- df[, paste0("t", 1:T_horizon), drop = FALSE]

# Safety check
stopifnot(ncol(t_cols) == T_horizon)

# Computing the fitted MLE probability for each distinct word/path
df$phat_mle <- apply(t_cols, 1, path_prob_homogeneous, pi = pi_full, A = A_hat)

# empirical word proportion
df$empirical <- df$count / sum(df$count)

# reconstruct the word, removing padding spaces
df$word <- apply(t_cols, 1, function(x) paste0(x[x != " "], collapse = ""))

# We have the following interpreation: Rank 1 = most frequent / largest probability
df$rank_obs <- rank(-df$count, ties.method = "average")
df$rank_mle <- rank(-df$phat_mle, ties.method = "average")

# positive: word ranks higher in the observations than under the MLE
# negative: word ranks higher under the MLE than in the observations
df$rank_diff <- df$rank_mle - df$rank_obs

# avoiding extremely rare words dominating the table
df_check <- subset(df, count >= 5)

# Words more prominent in the observations than under the MLE
obs_higher <- df_check[order(-df_check$rank_diff),
                       c("word", "count", "empirical", "phat_mle",
                         "rank_obs", "rank_mle", "rank_diff")]

# Words more prominent under the MLE than in the observations
mle_higher <- df_check[order(df_check$rank_diff),
                       c("word", "count", "empirical", "phat_mle",
                         "rank_obs", "rank_mle", "rank_diff")]

# Display the largest discrepancies for both directions, as suggested by the reviewer. 
head(obs_higher, 10)
head(mle_higher, 10)
```
 
##Finally, making a succint table for the paper: 

```{r}

outlier_table <- rbind(
  data.frame(direction = "Higher in observations", head(obs_higher, 3)),
  data.frame(direction = "Higher under MLE", head(mle_higher, 3))
)

# Rounding numerical columns for readability sake
outlier_table$empirical <- signif(outlier_table$empirical, 4)
outlier_table$phat_mle <- signif(outlier_table$phat_mle, 4)
outlier_table$rank_obs <- round(outlier_table$rank_obs, 1)
outlier_table$rank_mle <- round(outlier_table$rank_mle, 1)
outlier_table$rank_diff <- round(outlier_table$rank_diff, 1)

outlier_table
```
Completed code. June 2026. 


